---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: html_document
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Abstract
========
Background
----------
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review. The purpose of these is to identify opportunities for improvement (OFI), which are necessary in order to implement corrective actions. Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods
-------
This was a registry-based study using all patients between 2014 and 2022 from the Karolinska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference. Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: ...

Results
-------

Conclusion
----------




Introduction
============
Trauma is a leading cause of mortality and morbidity worldwide. According to WHO, injury-related deaths constitute 4% of all deaths and take the lives of 4,4 million people yearly worldwide - more than HIV/AIDS, malaria, maternal mortality and tuberculosis combined. The top 3 of 5 causes of death in people aged 5-29 are injury-related [1,2]. Trauma accounts for one-tenth of disability-adjusted life-years (DALYs), most of the burden in low and middle income countries (LMIC) [3, 4]. Trauma costs approximately 3% of GDP yearly globally [5].

A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review. The purpose of these is to identify opportunities for improvement (OFI). Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner [6]. Albaaj et al. recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna. The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [7]. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. 

There are many studies on preventable deaths in trauma patients treated in the ICU both regarding to the medical aspect as well as the human aspect. The primary medical causes of death after trauma are haemorrhage, sepsis/multiorgan failure and central nervous system injury. Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [8]. The most prominent OFI in regards to haemorrhage was decision-making compared to errors in technical skill. The OFIs frequently involved the decision between radiology, surgery and further investigation [9]. 

The University of Pennsylvania ICUs uses telemedicine and telemonitoring to review critical patient events. The areas of improvement they identified were: team dynamics and communication, use of best practices, increased and standardized access to emergency resources and improvement of procedural technique. They recommend other ICUs with telemedicine capacity to adopt a similar strategy in order to present OFIs and educational discussions [10]. Another study looked at how critical information was handled in the ICU and found that laboratory values and test results were the most frequently lost items. This points at the importance of communication and handling of critical information at handoffs of care [11]. From the studies mentioned above some OFIs related to the ICU can be identified: decision-making, access to resources and communication; however, there is still much left to discover. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.


Methods 
=======
Study design
------------
?

Setting
-------
The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [12], which means that all patients in Stockholm with severe or suspected severe trauma are sent there. The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [13]. All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses. The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for better organization or more senior members taking part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database. 

Setting
-------
The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], and is designated to receive all patients in Stockholm with major or suspected major trauma. The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6]. All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses. The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for better organization or more senior members taking part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database. 

Participants
------------
The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and June 1, 2022 and who were also admitted to the ICU. We excluded patients who were younger than 15 years and patients who were dead on arrival.

Variables and data sources/measurements
---------------------------------------

### Study outcome
The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors...

### Patient and process factors 
The factors were chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry. The categorical factors were sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated. The continuous factors age and Injury Severity Score (ISS) were also included. Working hours was defined as between 8.00 a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.

Bias 
----
We used R for all statistical analyses []. A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score... We present sample characteristics using descriptive statistics. The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database. 

Study size 
----------

<!-- Read up on the sample size requirements of logistic regression. You will come across for example the 10 events per variable rule, which is sometimes used as a rule of thumb to decide the size of the model in comparison to available data. You can also read up on the concept of power. -->

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------
*senare*

<!-- Age, ISS, time from arrival to CT, systolic blood pressure, respiratory rate and GCS score can all be considered quantitative variables. Describe how they were handled in the analyses. -->

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------
Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI. Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level. 

<!-- 

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed 

(d) Cohort study: If applicable, explain how loss to follow-up was addressed 

(e) Describe any sensitivity analyses
-->




Results
=======

<!-- 

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->



```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```




Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->





Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->




References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->
