---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: html_document
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Abstract
========
Background
----------
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions. Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods *redigera*
-------
This was a registry-based study using all patients between 2014 and 2022 from the Karoliniska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference. 

Results
-------

Conclusion
----------




Introduction
============
Trauma is a leading cause of mortality and morbidity worldwide being the leading cause of death among people under 44 years old in the United States and the fourth leading cause of death in the elderly [1]. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions. Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner rather than only subjective methods where individual cases are reviewed [2]. Examples of opportunities for improvement include lack of resources and management errors. 

Albaaj et al. recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna. The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [3]. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. 

According to a study on the Korean trauma registry the primary cause of death after trauma was haemorrhage, followed by sepsis/multiorgan failure and central nervous system injury. Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [4]. These are examples of factors that could be important in care on a higher hospital level such as the ICU. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.



```{r}

@ARTICLE{1,
  title     = "Short-term versus long-term trauma mortality: A systematic
               review",
  author    = "Frydrych, Lynn M and Keeney-Bonthrone, Toby P and Gwinn,
               Elizabeth and Wakam, Glenn K and Anderson, Maia S and Delano,
               Matthew J",
  journal   = "J. Trauma Acute Care Surg.",
  publisher = "Ovid Technologies (Wolters Kluwer Health)",
  volume    =  87,
  number    =  4,
  pages     = "990--997",
  month     =  oct,
  year      =  2019,
  language  = "en"
}

#https://pubmed.ncbi.nlm.nih.gov/31589196/
#Article on the evolution of long-term mortality in trauma patients as compared to the improved mortality numbers short-term


@ARTICLE{2,
  title    = "Identification of preventable trauma deaths: confounded
              inquiries?",
  author   = "Wilson, D S and McElligott, J and Fielding, L P",
  abstract = "The published evaluation of methods for identifying preventable
              trauma deaths contains many unstudied confounding factors. To
              investigate the reliability of methods for identifying such
              preventable deaths, we compared three consensus systems using
              separate five-member general review panels assessing 20
              non-central nervous system fatalities: panel A, independent
              judgments; panel B, discussion of all cases preceding individual
              judgments; and panel C, independent judgments followed by
              discussion and equivocal case reassignment. The Kappa concordance
              index was low for all methods (method A, 0.20; methods B and C,
              0.40). Of the 11 deaths judged preventable by at least one panel,
              only one death was judged preventable by all three panels.
              Consensus agreement (four of five assessors) was 20\% for panel
              A, 45\% for panel B, and 10\% for panel C (difference between
              panels B and C, p less than 0.03). In panel C, discussion
              affected the rate of equivocal case designation from 30\% to 5\%.
              Thus different consensus methods yielded different results. We
              conclude that individual case review can be severely flawed and
              therefore should not be used to measure institutional quality of
              patient care. We recommend that assessment of institutional
              performance should be based on objective evaluation methods,
              which require the study of patient population outcomes, rather
              than on subjective methods in which individual cases are
              reviewed.",
  journal  = "J. Trauma",
  volume   =  32,
  number   =  1,
  pages    = "45--51",
  month    =  jan,
  year     =  1992,
  language = "en"
}


#https://journals.lww.com/jtrauma/abstract/1992/01000/identification_of_preventable_trauma_deaths_.10.aspx

@ARTICLE{3,
  title    = "Patient and process factors associated with opportunities for
              improvement in trauma care: a registry-based study",
  author   = "Albaaj, Hussein and Attergrim, Jonatan and Str{\"o}mmer, Lovisa
              and Brattstr{\"o}m, Olof and Jacobsson, Martin and Wihlke,
              Gunilla and V{\"a}sterbo, Liselott and Joneborg, Elias and Gerdin
              W{\"a}rnberg, Martin",
  abstract = "BACKGROUND: Trauma is one of the leading causes of morbidity and
              mortality worldwide. Morbidity and mortality review of selected
              patient cases is used to improve the quality of trauma care by
              identifying opportunities for improvement (OFI). The aim of this
              study was to assess how patient and process factors are
              associated with OFI in trauma care. METHODS: We conducted a
              registry-based study using all patients between 2017 and 2021
              from the Karolinska University Hospital who had been reviewed
              regarding the presence of OFI as defined by a morbidity and
              mortality conference. We used bi- and multivariable logistic
              regression to assess the associations between the following
              patient and process factors and OFI: age, sex, respiratory rate,
              systolic blood pressure, Glasgow Coma Scale (GCS), Injury
              Severity Score (ISS), survival at 30 days, highest hospital care
              level, arrival on working hours, arrival on weekends, intubation
              status and time to first computed tomography (CT). RESULTS: OFI
              was identified in 300 (5.8\%) out of 5182 patients. Age, missing
              Glasgow Coma Scale, time to first CT, highest hospital care level
              and ISS were statistically significantly associated with OFI.
              CONCLUSION: Several patient and process factors were found to be
              associated with OFI, indicating that patients with moderate to
              severe trauma and those with delays to first CT are at the
              highest odds of OFI.",
  journal  = "Scand. J. Trauma Resusc. Emerg. Med.",
  volume   =  31,
  number   =  1,
  pages    = "87",
  month    =  nov,
  year     =  2023,
  keywords = "Acute surgery; Opportunities for improvement; Trauma; Trauma care",
  language = "en"
}

#Tidigare publicerad artikel av NOACs
#Albaaj, H. et al. Patient and process factors associated with opportunities for improvement in trauma care: a registry based study. Scandinavian Journal of Trauma, Resuscitation and Emergency Medicine (2023) 31:87.https://doi.org/10.1186/s13049-023-01157-y.


@ARTICLE{4,
  title    = "National assessment of opportunities for improvement in
              preventable trauma deaths: A mixed-methods study",
  author   = "Kwon, Junsik and Lee, Myeonggyun and Jung, Kyoungwon",
  abstract = "Trauma is a significant public health issue worldwide,
              particularly affecting economically active age groups. Quality
              management of trauma care at the national level is crucial to
              improve outcomes of major trauma. In Korea, a biennial nationwide
              survey on preventable trauma death rate is conducted. Based on
              the survey results, we analyzed opportunities for improving the
              trauma treatment process. Expert panels reviewed records of 8282
              and 8482 trauma-related deaths in 2017 and 2019, respectively,
              identifying 258 and 160 cases in each year as preventable deaths.
              Opportunities for improvement were categorized into prehospital,
              interhospital, and hospital stages. Hemorrhage was the primary
              cause of death, followed by sepsis/multiorgan failure and central
              nervous system injury. Delayed hemostatic procedures and
              transfusions were common areas for improvement in hospital stage.
              Interhospital transfers experienced significant delays in arrival
              time. This study emphasizes the need to enhance trauma care by
              refining treatment techniques, centralizing patients in
              specialized facilities, and implementing comprehensive reviews
              and performance improvements throughout the patient transfer
              system. The findings offer valuable insights for addressing
              trauma care improvement from both clinical and systemic
              perspectives.",
  journal  = "Healthcare (Basel)",
  volume   =  11,
  number   =  16,
  month    =  aug,
  year     =  2023,
  keywords = "mortality; patient transfer; quality improvement; treatment
              outcome; wounds and injuries",
  language = "en"
}


#https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10454732/pdf/healthcare-11-02291.pdf

```



Methods 
=======
Study design
------------
We conducted single-centre registry-based retrospective cohort study using data from the trauma registry and trauma care quality database at Karolinska University Hospital in Solna. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome of the selected patients was OFI identified by the multidisciplinary review board and then recorded in the trauma care quality database. From 2017 the trauma care quality database includes all patients in the trauma registry as well as the results of the mortality and morbidity review. This project will use these two databases and assess what types of OFI occurred in ICU patients and how different patient factors; such as age, sex, mechanism of injury and injury severity, are associated with OFI using logistic regression. A 5% significance level and 95% confidence levels will be used.


Setting
-------
The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], which means that all patients in Stockholm with severe or suspected severe trauma are sent there. The hospital has direct access to radiology, surgery, intensive care and consultants from relevant specialities [6]. All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses. The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for better organization or more senior members taking part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database. 


```{r}

@ARTICLE{5,
  title    = "Trauma center levels defined",
  author   = "Trauma Center Association of America",
  URL  = "https://www.traumacenters.org/page/TraumaCenterLevels#Goals",
  year     =  2014,
#Hur källhänvisar man webbsida?
 
@ARTICLE{6,
  title    = "Traumamanual",
  author   = "Karolinska Universitetssjukhuset Solna",
  year     =  2020,
#Include ISS? And audit filters?

```

Participants
------------
The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and *June 1, 2022*. *We excluded patients who were younger than 15 years and patients who were dead on arrival.*


Variables and data sources/measurements
---------------------------------------
Study outcome
The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors. *vilka OFIs ska läggas till för att analyseras?*

Patient and process factors 
The factors were chosen from the trauma registry, based on the locally used audit filters, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry. The categorical factors were *sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated*. The continuous factors age and Injury Severity Score (ISS) were also included. Working hours was defined as between 8.00 a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.

Statistical analysis 
Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level. The programming language used was R for all analyses []. The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database. 

*lägg till?*
A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above. We present sample characteristics using descriptive statistics. Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI. 

```{r}
# LÄNKA till http://www.rstudio.com/ 
```


Bias 
----
*senare*

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------
*senare*

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------
*senare*

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------
*senare*

<!-- 

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed 

(d) Cohort study: If applicable, explain how loss to follow-up was addressed 

(e) Describe any sensitivity analyses
-->





Results
=======

<!-- 

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->



```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```




Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->





Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->




References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->
